ST563 Final Project

### 2021-04-20 DRAFT.Rmd
mana edits

```{r good v bad}
good.wine <- subset(wine, quality > 5)
bad.wine <- subset(wine, quality <= 5)
summary(good.wine)
summary(bad.wine)
```
The two groups were similar in most of the variables. Two notable differences are that the wines classified as "good" had a lower mean value than "bad" wines for total sulfur dioxide. Good wines had a mean of 39, which was significantly lower than the mean of the "bad" wines at 54.6. Another difference to note is the difference in the mean for alcohol. The good wines had a mean alcohol content at 10.8, while the bad wines had a lower mean alcohol content around 9.9. There were also differences in fixed acidity, with the bad wines having a lower mean. There were also several variables that don't seem to change between the two groups including pH, residual sugar, chlorides, and density.

@@

# 1. full size decision tree
tree.class<-tree(class~.-quality,trn)
plot(tree.class); text(tree.class, pretty=0)
tree.class.pred<-predict(tree.class,tst,type='class')
tree.class.train_err<-mean(trn$class!=predict(tree.class,type='class'))
tree.class.test_err<-mean(tst$class!=tree.class.pred)
tree.class.train_err
tree.class.test_err
```
The resulting classification tree creates splits on 4 variables, Alcohol, Total Sulfur Dioxide, Sulphates, and Volatile Acidity. There are 10 terminal nodes. The tree has a test error rate of 0.297. This test rate is fairly low, but as mentioned before, if one conducts pruning on the tree, the error rate can decrease and a more accurate model is created.

Pruning is performed on the tree using cross validation, the tree is built many times with a different number of splits and the tree that produces the smallest error rate is selected. Cross validation on this model shows that the best model is the model built with 8 splits. The resulting pruned model will now have 8 terminal nodes and still uses the four variables in the un-pruned tree.    
```{r pruned tree}
# 2. pruned decision tree 
cv.tree.class<-cv.tree(tree.class,FUN=prune.misclass)
plot(cv.tree.class)
prune.tree.class<-prune.misclass(tree.class,best=8)
plot(prune.tree.class);text(prune.tree.class,pretty=0)
prune.tree.class.pred<-predict(prune.tree.class,tst,type='class')
prune.tree.class.train_err<-mean(trn$class!=predict(prune.tree.class,type='class'))
prune.tree.class.test_err<-mean(tst$class!=prune.tree.class.pred)
prune.tree.class.test_err
```
The pruned tree produces a test error rate to the original tree with a test MSE of 0.3. The pruned tree would be selected using the concept of Occam's Razor, where the simplest model is the best model choice.

Using a Random Forest tree can often improve upon the error rate of a simple classification tree. While the test MSE was small in the classification tree, the team believed the error rate could be decreases in using a more complex tree based method.

@@

The first thing that should be done in building a boosted tree is cross validation. Here the model is built many times with the number of predictors included from one to all predictors. The test MSE is evaluated for each and the smallest value is selected. A plot of these is included below, the tree that produced the smallest error rate was the tree built with two variables, so that was the tree selected through cross validation.  
```{r rf classification CV}
#----- classification -----
set.seed(702)
test.err <- double(11)
for (mtry in 1:11){
  fit <- randomForest(class~.-quality, trn,mtry=mtry)
  pred <- predict(fit,tst)
  test.err[mtry] <- mean(tst$class != pred)
}
matplot(1:mtry,test.err,pch=19,col='red',type='b',
        ylab='Mean Square Error')
```

```{r rf class models}
rf.class <- randomForest(class~.-quality,trn,mtry=2)
rf.class.pred <- predict(rf.class,tst)
rf.class.train_err <- mean(trn$class!=predict(rf.class))
rf.class.test_err <- mean(tst$class!=rf.class.pred)
rf.class.train_err
rf.class.test_err
```
Once the Random Forest tree with two variables was built, the final test MSE was calculated. This model produced a test MSE of 0.197, which is significantly lower than the test MSE for the classification model (0.3). The Random Forest model was selected as the best method for predicting if a wine were to fall into the "high" or "low" class when it comes to wine quality.

# Regression Methods
## Multiple Linear Regression
NEED CODE FROM YUN
```{r mlr}
#### REGRESSION ####
lm.reg <- lm(form.reg, data = trn)

@@

## Random Forest for Regression
As mentioned earlier in the classification section, Random Forest is an ensemble tree-based method, which creates multiple trees from bootstrap samples and averages the results from all the fitted trees. This model will be built exactly in the same way as the classification model, except the tree will not be predicting if the wine is a part of the good or bad class, but rather predicts the quality score associated with each wine. As before the model will first be built using cross validation to determine the number of variables that should be used in order to produce the smallest test MSE. Similar to the random forest classification tree, the cross validation selects the tree with 2 variables.
```{r rf regression}
#----- regression -----
test.err=double(11)
for (mtry in 1:11){
  fit<- randomForest(quality~.-class,trn,mtry=mtry)
  pred <- predict(fit,tst)
  test.err[mtry] <- sqrt(mean((tst$quality - pred)^2))
}
matplot(1:mtry,test.err,pch=19,col='red',type='b',
        ylab='Mean Square Error')

rf.reg<-randomForest(quality~.-class,trn,mtry=2)
rf.reg.pred<-predict(rf.reg,tst)
rf.reg.train_err<-sqrt(mean((trn$quality-predict(rf.reg))^2))
rf.reg.test_err<-sqrt(mean((tst$quality-rf.reg.pred)^2))
rf.reg.train_err
rf.reg.test_err
```  
The test MSE for the Regression Random Forest using 2 variables is 0.55. COMPARE THIS TO MLR WHEN WE GET YUN'S CODE.

I'm not sure we need these next 2 sections because most of the methods we are using do model selection. I don't have Yun's code, but I'm assuming he does some sort of variable selection, so this part seems extraneous. 
## Cross Validation
Again, our focus is on the classification accuracy of machine learning methods. In particular, we will be using traditional cross-validation methods to assess the accuracy, sensitivity, and specificity of each of the models. We split the data into a training and test data set in order to later evaluate the model's prediction accuracy or root mean squared error (RMSE). For this project we used a 80/20 split, training the data on the 80% and testing the trained models on the withheld 20%. We will then repeat this process over 5 folds of the data, averaging the results.

```{r CV}
# Specify the cross-validation method (5-fold CV)
trctrl <- trainControl(method = "cv", number = 5)
```

```{r class.select, out.height='4in'}
# Write preliminary model  
model.class.pre <- class ~ .

```{r reg.select, out.height='3.5in'}
# Create model and get summary output
lm.sub <- regsubsets(quality ~ ., data = wine[,-13], nvmax = 11)
smre <- lm.sub %>% summary()

```{r form}
# Write formula for regression models
form.reg <- quality ~ volatile.acidity + chlorides + pH + 
            total.sulfur.dioxide + sulphates + alcohol

@@

```{r create.df}
# Create data frame output for regression models
df.reg <- rbind.data.frame(lm.reg.rmse, rf.reg.rmse) %>% round(4)
df.reg$method <- reg
colnames(df.reg) <- c(LETTERS[1:2])
df.reg <- df.reg %>% select("B", "A")
colnames(df.reg) <- c("Regression Method", "Root MSE")

# Create data frame output for classification models
df.class <- rbind.data.frame(prune.tree.class.test_err, rf.class.test_err) %>% round(4)
colnames(df.class) <- c("Misclassification Rate")
rownames(df.class) <- c("Classification Model", "Random Forest")
df.class
```

@@

>>>>

```{r good v bad}
good.wine <- subset(wine, quality > 5)
bad.wine <- subset(wine, quality <= 5)
# summary(good.wine)
# summary(bad.wine)
```
The two groups were similar in most of the variables. Two notable differences are that the wines classified as "good" had a lower mean value than "bad" wines for total sulfur dioxide. Good wines had a mean of 39, which was significantly lower than the mean of the "bad" wines at 54.6. Another difference to note is the difference in the mean for alcohol. The good wines had a mean alcohol content at 10.8, while the bad wines had a lower mean alcohol content around 9.9. There were also differences in fixed acidity, with the bad wines having a lower mean. There were also several variables that don't seem to change between the two groups including pH, residual sugar, chlorides, and density.

@@

# 1. full size decision tree
tree.class <- tree(class ~ . - quality, trn)
plot(tree.class); text(tree.class, pretty = 0)
tree.class.pred <- predict(tree.class, tst, type = 'class')
tree.class.train_err <- mean(trn$class != predict(tree.class, type = 'class'))
tree.class.test_err <- mean(tst$class != tree.class.pred)
# tree.class.train_err
# tree.class.test_err
```

The resulting classification tree creates splits on 4 variables: Alcohol, Total Sulfur Dioxide, Sulphates, and Volatile Acidity. The tree has a test error rate of `r round(tree.class.test_err, 3)`. This test rate is fairly low, but as mentioned before, if one conducts pruning on the tree, the error rate can decrease and a more accurate model is created.

Pruning is performed on the tree using cross validation, the tree is built many times with a different number of splits and the tree that produces the smallest error rate is selected. Cross validation on this model shows that the best model is the model built with 8 splits. The resulting pruned model will now have 8 terminal nodes and still uses the four variables in the un-pruned tree.

```{r pruned tree}
# 2. pruned decision tree 
cv.tree.class <- cv.tree(tree.class, FUN = prune.misclass)
plot(cv.tree.class)
prune.tree.class <- prune.misclass(tree.class, best = 8)
plot(prune.tree.class); text(prune.tree.class, pretty = 0)
prune.tree.class.pred <- predict(prune.tree.class, tst, type = 'class')
prune.tree.class.train_err <- mean(trn$class != predict(prune.tree.class, type = 'class'))
prune.tree.class.test_err <- mean(tst$class != prune.tree.class.pred)
# prune.tree.class.test_err
```

The pruned tree produces a test error rate to the original tree with a misclassification error of `r prune.tree.class.test_err`.Using a Random Forest tree can often improve upon the error rate of a simple classification tree. While the test MSE was small in the classification tree, the team believed the error rate could be decreased using a more complex tree based method.

@@

The first thing that should be done in building a boosted tree is cross validation. Here the model is built many times with the number of predictors included from one to all predictors. The test MSE is evaluated for each and the smallest value is selected. A plot of these is included below, the tree that produced the smallest error rate was the tree built with two variables, so that was the tree selected through cross validation.

```{r rf classification CV}
#----- classification -----
set.seed(702)
test.err <- double(11)
for (mtry in 1:11){
  fit <- randomForest(class ~ . - quality, trn, mtry = mtry)
  pred <- predict(fit, tst)
  test.err[mtry] <- mean(tst$class != pred)
}

# Plot cross validation results
matplot(1:mtry, test.err, pch = 1, type = 'b',
        xlab = "mtry", ylab = 'Mean Square Error')
points(which.min(test.err), test.err[which.min(test.err)], pch = 8, col = "#CC0000")
```

```{r rf class models}
rf.class <- randomForest(class ~ . - quality, trn, mtry = 2)
rf.class.pred <- predict(rf.class,tst)
rf.class.train_err <- mean(trn$class != predict(rf.class))
rf.class.test_err <- mean(tst$class != rf.class.pred)
# rf.class.train_err
# rf.class.test_err
```

Once the Random Forest tree with two variables was built, the final test MSE was calculated. This model produced a test MSE of `r round(rf.class.test_err, 3)`, which is significantly lower than the test MSE for the basic classification tree (`r round(prune.tree.class.test_err`)). The random forest was selected as the best method for predicting if a wine were to fall into the "high" or "low" class when it comes to wine quality.

# Regression Methods
## Multiple Linear Regression
NEED CODE FROM YUN
```{r mlr}
#### REGRESSION ####
lm.reg <- lm(quality ~ . - class, data = trn)

@@

## Random Forest for Regression
As mentioned earlier in the classification section, Random Forest is an ensemble tree-based method, which creates multiple trees from bootstrap samples and averages the results from all the fitted trees. This model will be built exactly in the same way as the classification model, except the tree will not be predicting if the wine is a part of the good or bad class, but rather predicts the quality score associated with each wine. As before the model will first be built using cross validation to determine the number of variables that should be used in order to produce the smallest test MSE. Similar to the random forest classification tree, the cross validation selects the tree with 2 variables.

```{r rf regression}
#----- regression -----
# loop thru all the different values of mtry
test.err = double(11)
for (mtry in 1:11){
  fit<- randomForest(quality ~ . - class, trn, mtry = mtry)
  pred <- predict(fit, tst)
  test.err[mtry] <- sqrt(mean((tst$quality - pred)^2))
}

# Plot cross validation results
matplot(1:mtry, test.err, pch = 1, type = 'b', ylab = 'Mean Square Error', xlab = "mtry")
points(which.min(test.err), test.err[which.min(test.err)], pch = 8, col = "#CC0000")

# fit the random forest with the best results
rf.reg <- randomForest(quality ~ . - class, trn, mtry = 2)
rf.reg.pred <- predict(rf.reg, tst)
rf.reg.train_err <- sqrt(mean((trn$quality - predict(rf.reg))^2))
rf.reg.test_err <- sqrt(mean((tst$quality - rf.reg.pred)^2))
#rf.reg.train_err
#rf.reg.test_err
```  
The test MSE for the Regression Random Forest using 2 variables is 0.55. COMPARE THIS TO MLR WHEN WE GET YUN'S CODE.

@@

```{r create.df}
# Create data frame output for regression models
df.reg <- rbind.data.frame(lm.reg.rmse, rf.reg.test_err) %>% round(4)
df.reg$method <- reg
colnames(df.reg) <- c(LETTERS[1:2])
df.reg <- df.reg %>% select("B", "A")
colnames(df.reg) <- c("Regression Method", "Root MSE")

# Create data frame output for classification models
df.class <- rbind.data.frame(prune.tree.class.test_err, rf.class.test_err) %>% round(4)
colnames(df.class) <- c("Misclassification Rate")
rownames(df.class) <- c("Classification Tree", "Random Forest")
```

@@

# NOT NEEDED ???

I'm not sure we need these next 2 sections because most of the methods we are using do model selection. I don't have Yun's code, but I'm assuming he does some sort of variable selection, so this part seems extraneous.

## Cross Validation
Again, our focus is on the classification accuracy of machine learning methods. In particular, we will be using traditional cross-validation methods to assess the accuracy, sensitivity, and specificity of each of the models. We split the data into a training and test data set in order to later evaluate the model's prediction accuracy or root mean squared error (RMSE). For this project we used a 80/20 split, training the data on the 80% and testing the trained models on the withheld 20%. We will then repeat this process over 5 folds of the data, averaging the results.

```{r CV, eval = F}
# Specify the cross-validation method (5-fold CV)
trctrl <- trainControl(method = "cv", number = 5)
```

```{r class.select, out.height='4in', eval = F}
# Write preliminary model  
model.class.pre <- class ~ .

```{r reg.select, out.height='3.5in', eval = F}
# Create model and get summary output
lm.sub <- regsubsets(quality ~ ., data = wine[,-13], nvmax = 11)
smre <- lm.sub %>% summary()

```{r form, eval = F}
# Write formula for regression models
form.reg <- quality ~ volatile.acidity + chlorides + pH + 
            total.sulfur.dioxide + sulphates + alcohol
